Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-03 paper.tex
Update the recombination plot.

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
    \includegraphics{figures/ancestry-perf-expected}
\end{center}
\caption{\label{fig-ancestry-perf}
    \jkcomment{This is a temporary version of the figure pending the outcome
    of discussion in \#96}
    \textbf{TOP}
    Approximate run time for \msprime\ simulations,
    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black line is a quadratic fit separately in each panel.
    \textbf{BOTTOM}
    Time data from compared with
    equation~\eqref{eqn-hsw-bound} (``expected'').
    Left panel shows observed times plotted against expected number
    of operations; right panel shows the ratio of the two, plotted against $\rho$.
}
\end{figure}

@@

More visibly to the end user, we also have a high standard for documentation,
with precise, comprehensive, and cross-linked documentation
that is automatically built from the code base
and served through the website \url{https://tskit.dev}.
With the goal of lowering the entry barrier to new users,
we have invested significant effort in writing examples and introductions,
and making common tasks discoverable.
We also view contributions to documentation as equally important to the project
as writing code or designing methods---what use would it be
to write reliable, stable software if no-one used it?

@@

% Simulation software is historically a total shitshow

Simulation is a vital tool in population genetics, but the approach to
developing simulation software has historically been dysfunctional. A huge
proliferation of tools have been published (the references here are not
exhaustive) and only a small minority of these are actively developed and
maintained today. The ecosystem is highly fragmented, with numerous different
ways of specifying parameters and representing results, and there are
significant software quality issues at all stages. This is unsurprising, since
the majority of simulation software development is performed by students, who
are themselves learning to code, often without formal training in software
development. Although developing a simulator is certainly a learning
opportunity, it is also an opportunity to learn poor development practices. A
large number of unmaintained tools with mostly overlapping features represents
a profound waste of effort, and it is worth considering what could have been
achieved if even a small fraction of this effort had been invested in shared
community infrastructure.

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences [citations] that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
% I'm going out on a limb here - do folks agree?
This is a learning experience for many users,
allowing them to improve their software development skills in a
welcoming community environment and being exposed to
state-of-the-art infrastructure. We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high-quality
and sustainable in the long term.

\jkcomment{TODO either need another paragraph linking these up, or we just
 drop this last one?}

Recent breakthroughs in methods to infer genetic ancestry in recombining
organisms~\citep{kelleher2019inferring,
speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
have made it possible to estimate such ancestry at scale for
the first time~\citep{harris2019database,tang2019genealogy}.
Given such inferred ancestry, many exciting applications
become possible. For example, \cite{osmond2021estimating}
developed a method to estimate the location of genetic ancestors
based on inferred trees, and other uses are sure to follow.
Since the inferred genetic ancestry
becomes the input for other downstream inferences, it is vitally
important that these primary inferences are thoroughly validated,
with the detailed properties of the inferred ancestries
catalogued and understood.
As we have seen, \msprime\---through the succinct tree sequence
data structure and \tskit\ library---can provide simulated
genetic ancestry at any scale or degree of detail required, and should
provide a valuable source of ground-truth for evaluating
these methods.

>>>>

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for (A) small and (B) larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black lines are quadratic fits separately in each panel
    and sample size.
}
\end{figure}

@@

More visibly to the end user, we also have a high standard for documentation,
with precise, comprehensive, and cross-linked documentation
that is automatically built from the code base
and served through the website \texttt{https://tskit.dev}.
With the goal of lowering the entry barrier to new users,
we have invested significant effort in writing examples and introductions,
and making common tasks discoverable.
We also view contributions to documentation as equally important to the project
as writing code or designing methods---what use would it be
to write reliable, stable software if no-one used it?

@@

% We can simulate so much more now

This release makes \msprime\ able to produce simulations
with a great deal more biological realism, without a loss in speed and efficiency.
This ability will allow researchers to perform more reliable % different word here
power analyses, more robustly test new methods,
and carry out more reliable inferences,
without worrying that gene conversion or backmutation might be sabatoging
their applications to real data.
These advances have only been possible thanks to a distributed,
collaborative model of software development,
and the work of many people.
We hope the tool continues to be useful to many.

% Simulation software is historically a beautiful shitshow

Even though simulation has long been a vital tool in population genetics,
such collaborative software development has historically been uncommon.
A huge proliferation of tools have been published
(the references here are not exhaustive)
and only a small minority of these are actively developed and
maintained today. The ecosystem is highly fragmented, with numerous different
ways of specifying parameters and representing results, and there are
significant software quality issues at all stages. This is unsurprising, since
the majority of simulation software development is performed by students, who
are themselves learning to code, often without formal training in software
development.
Although this ``Haldane's sieve'' of software has produced many good tools
and enabled decades of research,
it also represents a missed opportunity to invest as a community
in shared infrastructure and mentorship in good software development practice.

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences \citet[OTHER CITATIONS]{siepel2019challenges}
that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
% I'm going out on a limb here - do folks agree?
% PLR- personally I have learned SO MUCH, and I'm passing it on to others
Software development in a welcoming community,
with mentorship by experienced developers,
is a hugely useul experience for many users,
The skills that contributors learn
can lead to greatly increased productivity in subsequent
work (e.g., through more reliable code and better debugging skills).
We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high-quality
and sustainable in the long term.

The tree sequence developed for \msprime\ allows users a view of not only
genetic variation but also the genetic ancestry that produced that variation.
Recent breakthroughs in methods to infer genetic ancestry in recombining
organisms~\citep{kelleher2019inferring,
speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
have made it possible to estimate such ancestry from real data at scale for
the first time~\citep{harris2019database,tang2019genealogy}.
Given such inferred ancestry, many exciting applications
become possible. For example, \cite{osmond2021estimating}
developed a method to estimate the location of genetic ancestors
based on inferred trees, and other uses are sure to follow.
Since the inferred genetic ancestry
becomes the input for other downstream inferences, it is vitally
important that these primary inferences are thoroughly validated,
with the detailed properties of the inferred ancestries
catalogued and understood.
\texttt{Msprime} will be an important tool for these inferences and validations,
and in this context the ability to interoperate
with other methods---particularly, forwards simulators---through
the succinct tree sequence data structure and \tskit\ library
will be vital.
